\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2024-03-18 docs/papers/sto/so.tex
FIXUP update figures of f and g nets

Cosmological $N$-body simulations evolve dark matter particles under the
gravitational force.
The dynamics of the particles is governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $- \vnabla \varphi$
follows
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
we then introduce SO force sharpening through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of
the wavevector components and all possible relevant features
$\vvartheta$.
The subscript $i\in\{1,2,3\}$ denotes the 3 principal axes of the
simulation box.

@@

\begin{figure*}
  \centering
  \includegraphics[width=.48\columnwidth]{net_f_a1_j3091776_e3000.pdf}
  \includegraphics[width=.48\columnwidth]{net_g_a1_j3091776_e3000.pdf}
  \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
  functions. The solid line and shaded region indicate the median and
  the [16, 84] percentiles of the 64 sampled configurations and
  cosmological parameters. $a=1$,
  mesh shape=1}
  \label{fig:func}
\end{figure*}

@@

\subsection{Benchmarks}
We compare the computational cost of \GADGET, original PM, PM-SONN, and
PM-SOSR.

The forward evaluation of the NNs is more computation and memory
expensive compared to the analytic expressions.
We time the computation of SO for one step using NNs and SR
respectively.

@@

We perform distributed data-parallel training using GPU devices hosted
on multiple computing nodes.
Each process is binded to a single GPU device, and the training data
samples are uniformly distributed to the processes.
In total, we use XXX \GPU\ GPUs, each accompanied with XXX CPU cores
and XXX GB CPU memory mainly for preloading the training data.
The order of data samples within each process is shuffled across
training epochs.
In each training step, a data sample is fed into the GPU from the CPU
memory where the preloading took place to save the repeated reading time
from the disk.

@@

We initialize the parameters of NNs in a way such that the output is
very close to one, i.e. no SO.
This is achieved by setting the kernel of the output layer to be very
small random values and the bias to be zero, whose return is then around
one after the exponential function.
The parameters are also initialized to be the same for all devices, and
hence remain the same during training given the same averaged gradients.

>>>>

Cosmological $N$-body simulations evolve dark matter particles under the
gravitational force.
The dynamics of the particles is governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $- \vnabla \varphi$
follows
%
\begin{equation}
\left[- \vnabla \varphi\right]_\vk = - i \vk \varphi(\vk) =
                  i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\label{force}
\end{equation}
%
which then gives the force field in configuration space after an inverse
Fourier transform.

In the PM simulation, the density field $\delta(\vx)$ is estimated by
scattering the particles onto the mesh, and then the force on each
particle is estimated by gathering the force field from the mesh.
Therefore, the accuracy of the gravitational force estimated for each
particle depends on the resolution of the mesh.
To improve the accuracy of the PM solver, we introduce SO force
sharpening by modifying the force equation \eqref{force} through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of
the wavevector components and all possible relevant features
$\vvartheta$.
The subscript $i\in\{1,2,3\}$ denotes the 3 principal axes of the
simulation box.

@@

\begin{figure*}
  \centering
  \includegraphics[width=.48\columnwidth]{net_f_a1_j3177874_e3000.pdf}
  \includegraphics[width=.48\columnwidth]{net_g_a1_j3177874_e3000.pdf}
  \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
  functions. The solid line and shaded region indicate the median and
  the [16, 84] percentiles of the 64 sampled configurations and
  cosmological parameters. $a=1$,
  mesh shape=1}
  \label{fig:func}
\end{figure*}

@@

\subsection{Benchmarks}
We compare the computational cost of \GADGET, original PM, PM-SONN, and
PM-SOSR.
In the generation of our training data, on average each \GADGET\
simulation takes about 2K CPU hours.

The forward evaluation of the NNs is more computation and memory
expensive compared to the analytic expressions.
We time the computation of SO for one step using NNs and SR
respectively, both after the just-in-time (JIT) compilation function
transformation in JAX.

@@

We perform distributed data-parallel training using GPU devices hosted
on multiple computing nodes.
With data parallelism, the training data samples are uniformly
distributed to the processes, and each process is binded to a single GPU
device.
In total, we use XXX \GPU\ GPUs, each accompanied with XXX CPU cores
and XXX GB CPU memory mainly for preloading the training data.
The order of data samples within each process is shuffled across
training epochs.
In each training step, a data sample is fed into the GPU from the CPU
memory where the preloading took place to save the repeated reading time
from the disk.

@@

We initialize the parameters of NNs in a way such that the output is
around one with slight deviations, i.e. roughly the original PM without
SO.
This is achieved by setting the kernel of the output layer to be very
small random values and the bias to be zero, whose return is then around
one after the exponential function.
The parameters are initialized to be the same for all devices, and hence
remain the same during training given the same averaged gradients.
